---
title: "Introduction to DuckDBGRanges"
author:
- name: Patrick Aboyoun
  email: aboyounp@gene.com
  affiliation: Genentech, Inc.
date: "Compiled: `r BiocStyle::doc_date()`; Modified: 6 July 2026"
package: DuckDBGRanges
vignette: >
  %\VignetteIndexEntry{1. Introduction to DuckDBGRanges}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      error = FALSE, warning = FALSE, message = FALSE)
```

# Introduction

Genomic workflows routinely produce interval sets that are awkward to hold in
memory: a gnomAD-scale variant catalogue is hundreds of millions of positions, a
multi-sample scATAC-seq study is millions of peaks, a comprehensive gene model is
millions of exons. Loading the whole set into an R session, alongside the
intermediate objects an analysis creates, is wasteful when a workflow only ever
touches a region, a chromosome, or the ranges passing a metadata filter.

`r Biocpkg("GenomicRanges")` is the standard Bioconductor representation for such
data, and its API (`seqnames()`, `start()`, `restrict()`, `findOverlaps()`)
is what most genomic code is written against. What varies is *where the data
lives*: an ordinary `GRanges` holds every range in memory.

`r Biocpkg("DuckDBGRanges")` provides a `GRanges` (and `GRangesList`) backed by
columnar **Parquet** queried through [DuckDB](https://duckdb.org). It exposes the
full `r Biocpkg("GenomicRanges")` API, but the coordinates stay **on disk** and
operations are recorded as lazy SQL queries: you can filter by region, restrict,
subset, and compose pipelines **without loading the ranges into memory**. The R
object stays a small, constant size (~175 KB) whether it fronts a thousand ranges
or a hundred million.

It builds on `r Biocpkg("DuckDBDataFrame")` (the tabular foundation of the
**BiocDuckDB** suite) and is designed for the *filter-first* shape of most genomic
work: use DuckDB to cut a large interval set down to the ranges of interest, then
materialize that manageable subset to an ordinary `GRanges` for the interval-tree
operations R does best.

This vignette is a practical introduction. Two companion vignettes go further:

- *Benchmarking DuckDBGRanges* measures what the on-disk backend costs and buys,
  against in-memory `GRanges`, on scATAC-seq and variant-scale data.
- *Design and extension of DuckDBGRanges* documents the class structure and SQL
  translation for developers.

## Installation

```{r install, eval=FALSE}
if (!require("BiocManager"))
    install.packages("BiocManager")
BiocManager::install("DuckDBGRanges")
```

```{r load}
library(DuckDBGRanges)
library(GenomicRanges)
```

# Quick start

A `DuckDBGRanges` is backed by a Parquet file with columns for the genomic
coordinates. For a small example we build a `GRanges`, write it to Parquet with
`r CRANpkg("arrow")`, and open it as a `DuckDBGRanges`.

```{r quickstart-data}
library(arrow)

gr <- GRanges(
    seqnames = rep(c("chr1", "chr2"), c(3, 2)),
    ranges = IRanges(start = c(100, 200, 300, 150, 250),
                     end   = c(150, 275, 360, 230, 295)),
    strand  = c("+", "-", "+", "-", "+"),
    gene_id = paste0("GENE", 1:5),
    score   = c(10, 20, 15, 25, 18))

gr_df <- as.data.frame(gr)
gr_df$range_id <- paste0("range", seq_len(nrow(gr_df)))
gr_path <- tempfile(fileext = ".parquet")
write_parquet(gr_df, gr_path)
```

We construct the object by naming the coordinate columns; extra columns become
metadata columns (`mcols`). The `keycol` argument names the column that identifies
each range.

```{r quickstart-construct}
gr_ddb <- DuckDBGRanges(gr_path,
                        seqnames = "seqnames", start = "start",
                        end = "end", strand = "strand",
                        keycol = list(range_id = gr_df$range_id),
                        mcols = c("gene_id", "score"))
gr_ddb
```

It looks and behaves like a `GRanges`, but the ranges stay on disk:

```{r quickstart-ops}
length(gr_ddb)
seqnames(gr_ddb)
```

# Working with a DuckDBGRanges

## Accessors

All the standard `GRanges` accessors work and return the same kinds of values,
queried from disk on demand:

```{r accessors}
start(gr_ddb)
width(gr_ddb)
ranges(gr_ddb)
gr_ddb$gene_id
```

## Subsetting

Ranges can be selected by index, by a logical condition on a coordinate or
metadata column (which becomes a SQL `WHERE` clause), or by overlap with a query
region:

```{r subset}
gr_ddb[1:3]
gr_ddb[seqnames(gr_ddb) == "chr1"]
gr_ddb[gr_ddb$score > 15]
gr_ddb[gr_ddb %over% GRanges("chr1:150-250")]
```

## Range operations

The intra- and inter-range operations of `r Biocpkg("GenomicRanges")` are
implemented as SQL, so the familiar functions work directly:

```{r range-ops}
shift(gr_ddb, 50L)
restrict(gr_ddb, start = 200L, end = 300L)
reduce(gr_ddb)
```

`coverage()` computes per-base depth the same way, entirely in SQL, and returns
an ordinary (small) `RleList`, one run-length-encoded vector per seqname:

```{r coverage}
coverage(gr_ddb)
```

## Filter, then materialize

The idiomatic workflow uses DuckDB to shrink a large set to the ranges of
interest, then coerces to an ordinary `GRanges` for the interval-tree operations
(`findOverlaps()`, `nearest()`) that in-memory algorithms do best:

```{r hybrid}
## lazy, on-disk filter down to a manageable subset...
sub <- gr_ddb[seqnames(gr_ddb) == "chr1" & gr_ddb$score > 12]

## ...then materialize the small result for overlap work
sub_gr <- as(sub, "GRanges")
class(sub_gr)
findOverlaps(sub_gr, GRanges("chr1:100-320"))
```

# Grouped ranges: DuckDBGRangesList

Grouped features (a transcript's exons, a gene's peaks) are held by a
`DuckDBGRangesList`, which stores each group's coordinates as DuckDB `LIST[]`
columns, one row per group, and behaves like a `GRangesList`:

```{r grangeslist}
grl_data <- data.frame(
    transcript_id = c("ENST001", "ENST002"),
    gene_id = c("GENE1", "GENE2"),
    seqnames = I(list(rep("chr1", 3), rep("chr2", 2))),
    start  = I(list(c(100L, 200L, 300L), c(150L, 250L))),
    width  = I(list(c(50L, 75L, 60L),   c(80L, 45L))),
    strand = I(list(c("+", "-", "+"),   c("-", "+"))))
grl_path <- tempfile(fileext = ".parquet")
write_parquet(grl_data, grl_path)

grl_ddb <- DuckDBGRangesList(grl_path,
                             seqnames = "seqnames", start = "start",
                             width = "width", strand = "strand",
                             mcols = "gene_id",
                             keycol = list(transcript_id = grl_data$transcript_id))
elementNROWS(grl_ddb)
grl_ddb[[1]]
```

# When to use DuckDBGRanges

A good fit when the interval set is **larger than memory** (or you want to keep
memory free), when the workload is **filter-heavy**, restricting to regions,
chromosomes, or metadata thresholds before the expensive step, or when the data
already lives on disk as **Parquet** that other tools (Python, Julia, cloud query
engines) can read. Because the R object is a constant ~175 KB, a `DuckDBGRanges`
also lets you *reference* a huge annotation without paying for it until a query
touches it.

An in-memory `GRanges` remains preferable when the data fits comfortably in RAM,
and for **overlap enumeration** (`findOverlaps()`, `subsetByOverlaps()`),
where the in-memory interval trees are hard to beat. (Filtering and
`distanceToNearest()`, by contrast, are faster on the DuckDB backend, see the
benchmarking vignette.) The two compose well: the *filter, then materialize*
pattern above uses each where it is strongest.

For a quantitative comparison see *Benchmarking DuckDBGRanges*; for the class
structure and SQL translation see *Design and extension of DuckDBGRanges*.

# Session information

```{r sessioninfo}
sessionInfo()
```
