---
title: "Design and extension of DuckDBGRanges"
author:
- name: Patrick Aboyoun
  email: aboyounp@gene.com
  affiliation: Genentech, Inc.
date: "Compiled: `r BiocStyle::doc_date()`; Modified: 6 July 2026"
package: DuckDBGRanges
vignette: >
  %\VignetteIndexEntry{3. Design and extension of DuckDBGRanges}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      error = FALSE, warning = FALSE, message = FALSE)
```

# Scope

This vignette is for developers: it documents how `r Biocpkg("DuckDBGRanges")`
represents genomic ranges on top of `r Biocpkg("DuckDBDataFrame")`, how range
operations become SQL, and how the classes are structured so you can maintain or
extend the package. For day-to-day use see *Introduction to DuckDBGRanges*; for
performance see *Benchmarking DuckDBGRanges*.

```{r load}
library(DuckDBGRanges)
library(GenomicRanges)
```

# Architecture

`r Biocpkg("DuckDBGRanges")` layers two classes on the
`r Biocpkg("DuckDBDataFrame")` foundation, each extending its in-memory
`r Biocpkg("GenomicRanges")` counterpart:

| Class | Extends | Role |
|-------|---------|------|
| `DuckDBGRanges` | `GRanges` | genomic intervals with a DuckDB backend |
| `DuckDBGRangesList` | `GRangesList` | grouped features (transcripts, per-gene peaks) |

Each holds its coordinates in a `DuckDBDataFrame` (from
`r Biocpkg("DuckDBDataFrame")`) describing a Parquet file, rather than in memory.
Because a `DuckDBGRanges` *is* a `GRanges`, the full `r Biocpkg("GenomicRanges")`
API applies unchanged; the package overrides the operations that benefit from
running as SQL, and falls back to materialization for the rest.

```{r classes}
showClass("DuckDBGRanges")
```

## Slots

A `DuckDBGRanges` carries three slots:

- **`frame`**: a `DuckDBDataFrame` with the coordinate columns (see [The five
  required columns]). Extra columns become metadata columns.
- **`seqinfo`**: a `Seqinfo` with sequence lengths and genome build, kept
  materialized (it is small and needed for validation).
- **`elementMetadata`**: an optional `DuckDBDataFrame` of metadata columns
  (`mcols`) sharing the same backend as `frame`.

`DuckDBGRangesList` has the same slots, but its coordinate columns are DuckDB
`LIST[]` columns (one row per group) and its `elementMetadata` is *per group*
rather than per range (see [Grouped ranges as LIST columns]).

# The five required columns

Every `DuckDBGRanges` resolves to five coordinate columns, mirroring the
components of a `GRanges`:

| Column | Meaning |
|--------|---------|
| `seqnames` | sequence/chromosome name (`"chr1"`, ...) |
| `start` | start position (1-based, inclusive) |
| `end` | end position (1-based, inclusive) |
| `width` | interval width (`end - start + 1`) |
| `strand` | `"+"`, `"-"`, or `"*"` |

The constructor names the columns backing these components; anything else named in
`mcols=` becomes a metadata column. Only the columns an operation actually needs
are ever read from the Parquet file.

```{r construct}
gr <- GRanges(c("chr1:100-150:+", "chr1:200-275:-", "chr2:300-360:+"),
              gene_id = paste0("GENE", 1:3))
gr_df <- as.data.frame(gr)
gr_df$range_id <- paste0("range", seq_len(nrow(gr_df)))
gr_path <- tempfile(fileext = ".parquet"); arrow::write_parquet(gr_df, gr_path)

gr_ddb <- DuckDBGRanges(gr_path,
                        seqnames = "seqnames", start = "start",
                        end = "end", strand = "strand",
                        keycol = list(range_id = gr_df$range_id),
                        mcols = "gene_id")
gr_ddb
```

# From range operations to SQL

The performance story rests on one idea: rather than materialize the ranges and
run the in-memory algorithm, `r Biocpkg("DuckDBGRanges")` translates the
operations that map cleanly onto set-based queries into SQL on the underlying
`DuckDBDataFrame`, so DuckDB does the scan and the arithmetic and only the result
crosses back into R.

**Coordinate filters** (`restrict()`, logical subsetting) become `WHERE`
clauses, which DuckDB pushes into the Parquet reader, so a region filter reads
only the matching row groups instead of scanning everything:

```sql
-- restrict(x, start = s, end = e)  becomes
SELECT * FROM ranges WHERE start >= s AND end <= e
```

**Overlap joins** (`findOverlaps()` and friends) become a self- or cross-join on
the interval-overlap predicate:

```sql
SELECT a.id, b.id
FROM ranges_a a JOIN ranges_b b
  ON a.seqnames = b.seqnames
  AND a.start <= b.end AND a.end >= b.start
  AND (a.strand = b.strand OR a.strand = '*' OR b.strand = '*')
```

**Distance** becomes a `CASE` expression over the gap between two intervals:

```sql
SELECT CASE
    WHEN a.end < b.start THEN b.start - a.end
    WHEN b.end < a.start THEN a.start - b.end
    ELSE 0
END AS distance
FROM ranges_a a, ranges_b b
```

Filters compose lazily, a chain of `restrict()` and logical subsets becomes a
single query, evaluated only when results are pulled. Operations without a natural
set-based form fall back to materializing the (already filtered) ranges to an
in-memory `GRanges`, which is correct for any input and cheap once the data has
been cut down. *Benchmarking DuckDBGRanges* shows which operations win in SQL and
which are better left to the in-memory interval-tree implementations.

# Grouped ranges as LIST columns

`DuckDBGRangesList` stores grouped features using DuckDB `LIST[]` columns: each
row is one group, and the coordinate columns hold lists of per-range values.

```
transcript_id | seqnames        | start        | width       | strand
ENST001       | [chr1,chr1,chr1]| [100,200,300]| [50,75,60]  | [+,-,+]
ENST002       | [chr2,chr2]     | [150,250]    | [80,45]     | [-,+]
```

This is compact (one row per group instead of one per range), matches the
biological grouping directly, and lets DuckDB's list functions operate on groups
without unnesting. `elementNROWS()` reads the list lengths; extracting a group
unnests it back to a `GRanges`.

```{r grl}
grl_data <- data.frame(
    transcript_id = c("ENST001", "ENST002"),
    seqnames = I(list(rep("chr1", 3), rep("chr2", 2))),
    start  = I(list(c(100L, 200L, 300L), c(150L, 250L))),
    width  = I(list(c(50L, 75L, 60L),   c(80L, 45L))),
    strand = I(list(c("+", "-", "+"),   c("-", "+"))))
grl_path <- tempfile(fileext = ".parquet"); arrow::write_parquet(grl_data, grl_path)

grl_ddb <- DuckDBGRangesList(grl_path,
                             seqnames = "seqnames", start = "start",
                             width = "width", strand = "strand",
                             keycol = list(transcript_id = grl_data$transcript_id))
elementNROWS(grl_ddb)
```

# Materialization and interoperability

A `DuckDBGRanges` coerces to an ordinary `GRanges` with `as(x, "GRanges")`, which
runs the accumulated query and reads the result into memory, the *materialize*
half of the *filter, then materialize* pattern. Coercion is how overlap-heavy
work is handed off to the in-memory interval-tree algorithms.

```{r materialize}
as(gr_ddb[seqnames(gr_ddb) == "chr1"], "GRanges")
```

Because the on-disk representation is plain Parquet, the same files are readable
outside R (Python, Julia, DuckDB itself, cloud query engines), the property the
`r Biocpkg("BiocDuckDB")` package relies on for its language-neutral on-disk
experiment format. A new range-bearing backend mostly needs to define how its
data maps onto the five coordinate columns and which operations to push into SQL;
everything else follows from the `r Biocpkg("GenomicRanges")` contract.

# Session information

```{r sessioninfo}
sessionInfo()
```
