---
title: "Architecture of the DuckDBGRanges Package"
author:
- name: Patrick Aboyoun
  email: aboyounp@gene.com
  affiliation: Genentech, Inc.
date: "Compiled: `r format(Sys.time(), '%B %d, %Y')`"
package: DuckDBGRanges
vignette: >
  %\VignetteIndexEntry{Architecture of the DuckDBGRanges Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_depth: 4
---

```{r setup, echo=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Introduction

The `r Biocpkg("DuckDBGRanges")` package provides DuckDB-backed implementations
of `r Biocpkg("GenomicRanges")` data structures for efficient, out-of-memory
operations on large genomic interval datasets. It builds upon the
`r Biocpkg("DuckDBDataFrame")` foundation to deliver genomic range-specific
functionality with SQL-optimized range operations.

## Position in the Bioconductor Ecosystem

`r Biocpkg("DuckDBGRanges")` extends the GenomicRanges framework with DuckDB
backend support:

```
DuckDBDataFrame (foundation: tabular data + SQL)
    ↓
DuckDBGRanges (genomic intervals + range operations)
    ↓
BiocDuckDB (high-level I/O + genomic workflows)
```

| Class | Extends | Purpose |
|-------|---------|---------|
| *DuckDBGRanges* | *GenomicRanges* | Genomic intervals with DuckDB backend |
| *DuckDBGRangesList* | *GRangesList* | Grouped genomic features (e.g., transcripts) |

The package is designed for genomic analyses requiring:
- **Large variant databases** (millions of positions)
- **Comprehensive gene models** (millions of exons)
- **scATAC-seq peaks** (hundreds of thousands of accessible regions)
- **Genome-wide annotations** (regulatory elements, repeats, etc.)

## Installation

```{r install, eval=FALSE}
if (!require("BiocManager"))
    install.packages("BiocManager")
BiocManager::install("DuckDBGRanges")
```

```{r load-package}
library(DuckDBGRanges)
library(DuckDBDataFrame)  # For examples
library(GenomicRanges)    # For comparison
```


# Core Design Principles

## GenomicRanges Integration

`r Biocpkg("DuckDBGRanges")` implements the full *GenomicRanges* API while
keeping data on disk in Parquet format. This provides:

1. **Familiar interface**: All standard GRanges operations work
2. **SQL optimization**: Range operations translate to efficient SQL
3. **Lazy evaluation**: Operations build queries without immediate execution
4. **Memory efficiency**: Only load data when needed

### The Five Required Columns

Every *DuckDBGRanges* object stores five essential columns:

- **seqnames**: Chromosome/sequence names (e.g., "chr1", "chr2")
- **start**: Start positions (1-based, inclusive)
- **end**: End positions (1-based, inclusive)
- **width**: Interval widths (end - start + 1)
- **strand**: Strand information (+, -, or *)

These are stored in a *DuckDBDataFrame* and accessed via SQL.


## SQL-Optimized Range Operations

Key range operations translate to efficient SQL:

### findOverlaps()
```sql
SELECT a.id, b.id
FROM ranges_a a JOIN ranges_b b
  ON a.seqnames = b.seqnames
  AND a.strand = b.strand
  AND a.start <= b.end
  AND a.end >= b.start
```

### distance()
```sql
SELECT
  CASE
    WHEN a.end < b.start THEN b.start - a.end
    WHEN b.end < a.start THEN a.start - b.end
    ELSE 0
  END as distance
FROM ranges_a a, ranges_b b
```

This SQL translation enables DuckDB's query optimizer to use indexes and
parallel execution.


## LIST[] Columns for Grouped Ranges

*DuckDBGRangesList* uses DuckDB's `LIST[]` columns to store grouped ranges
efficiently. Each row represents a group (e.g., a transcript), and the
genomic coordinates are stored as lists:

```
transcript_id | seqnames      | start         | end           | strand
ENST001       | ["chr1", ...] | [100, 200]    | [150, 250]    | ["+", "+"]
ENST002       | ["chr2", ...] | [300, 400]    | [350, 450]    | ["-", "-"]
```

This representation is:
- **Compact**: One row per group instead of one per range
- **Efficient**: DuckDB's LIST[] operations are optimized
- **Natural**: Matches the biological concept of grouped features


# The DuckDBGRanges Class

## Class Structure

*DuckDBGRanges* extends *GenomicRanges* and contains a *DuckDBDataFrame*:

```{r DuckDBGRanges-class}
showClass("DuckDBGRanges")
```

### Slots

**frame** (*DuckDBDataFrame*)
: A DuckDBDataFrame with exactly five columns: seqnames, start, end, width,
  and strand. Additional columns become metadata columns (mcols).

**seqinfo** (*Seqinfo*)
: Sequence information including chromosome lengths and genome build.

**elementMetadata** (*DuckDBDataFrame* or NULL)
: Metadata columns (mcols) stored as a separate DuckDBDataFrame sharing the
  same backend as frame.


## Construction

### From Parquet Files

```{r DuckDBGRanges-construct}
# Create example genomic ranges
gr <- GRanges(
    seqnames = rep(c("chr1", "chr2"), c(3, 2)),
    ranges = IRanges(start = c(100, 200, 300, 150, 250),
                     end = c(150, 275, 360, 230, 295)),
    strand = c("+", "-", "+", "-", "+"),
    gene_id = paste0("GENE", 1:5),
    score = c(10, 20, 15, 25, 18)
)

# Write to Parquet
gr_df <- as.data.frame(gr)
gr_df$range_id <- paste0("range", 1:nrow(gr_df))
gr_path <- tempfile(fileext = ".parquet")
arrow::write_parquet(gr_df, gr_path)

# Create DuckDBGRanges
gr_ddb <- DuckDBGRanges(gr_path,
                        seqnames = "seqnames",
                        start = "start",
                        end = "end",
                        strand = "strand",
                        keycol = list(range_id = gr_df$range_id),
                        mcols = c("gene_id", "score"))
gr_ddb
```

### From Existing GRanges

```{r DuckDBGRanges-from-gr, eval=FALSE}
# Convert in-memory GRanges to DuckDB-backed
gr_path2 <- tempfile(fileext = ".parquet")
arrow::write_parquet(as.data.frame(gr), gr_path2)
gr_ddb2 <- DuckDBGRanges(gr_path2, ...)
```


## Accessors

*DuckDBGRanges* provides all standard GRanges accessors:

### Genomic Coordinates

```{r DuckDBGRanges-coords}
# Sequence names
seqnames(gr_ddb)

# Start positions
start(gr_ddb)

# End positions
end(gr_ddb)

# Widths
width(gr_ddb)

# Strand
strand(gr_ddb)

# Combined ranges
ranges(gr_ddb)
```

### Metadata

```{r DuckDBGRanges-metadata}
# Metadata columns
mcols(gr_ddb)

# Individual metadata column
gr_ddb$gene_id

# Sequence information
seqinfo(gr_ddb)
```

### Length and Names

```{r DuckDBGRanges-length}
# Number of ranges
length(gr_ddb)

# Range names (if any)
names(gr_ddb)
```


## Subsetting

*DuckDBGRanges* supports all standard subsetting operations:

### By Index

```{r DuckDBGRanges-subset-index}
# First 3 ranges
gr_ddb[1:3]

# Specific ranges
gr_ddb[c(1, 3, 5)]
```

### By Logical Vector

```{r DuckDBGRanges-subset-logical}
# Ranges on chr1
gr_ddb[seqnames(gr_ddb) == "chr1"]

# Ranges with score > 15
gr_ddb[gr_ddb$score > 15]

# Positive strand ranges
gr_ddb[strand(gr_ddb) == "+"]
```

### By Overlaps

```{r DuckDBGRanges-subset-overlaps}
# Define query region
query <- GRanges("chr1:150-250")

# Subset to overlapping ranges
gr_ddb[gr_ddb %over% query]
```


# Range Operations

## Intra-Range Operations

Operations that modify ranges in place:

### shift()

```{r shift}
# Shift all ranges by 50bp
shifted <- shift(gr_ddb, 50)
start(shifted)
```

### resize()

```{r resize}
# Resize to 100bp from start
resized <- resize(gr_ddb, width = 100, fix = "start")
width(resized)
```

### flank()

```{r flank}
# Get 50bp upstream flanks
upstream <- flank(gr_ddb, width = 50, start = TRUE)
upstream
```

### promoters()

```{r promoters}
# Get promoter regions (2kb upstream, 200bp downstream)
proms <- promoters(gr_ddb, upstream = 2000, downstream = 200)
width(proms)
```

### restrict()

```{r restrict}
# Restrict to chr1:100-300
restricted <- restrict(gr_ddb, start = 100, end = 300)
restricted
```

### narrow()

```{r narrow}
# Narrow to middle 50bp
narrowed <- narrow(gr_ddb, start = 25, end = -25)
width(narrowed)
```


## Inter-Range Operations

Operations between multiple ranges:

### reduce()

```{r reduce}
# Merge overlapping ranges
reduced <- reduce(gr_ddb)
reduced
```

### gaps()

```{r gaps}
# Find gaps between ranges
gaps_gr <- gaps(gr_ddb)
gaps_gr
```

### disjoin()

```{r disjoin}
# Create disjoint ranges
disjoined <- disjoin(gr_ddb)
disjoined
```


## Between-Range Operations

Operations between two GRanges objects:

### findOverlaps()

The most important operation for genomic analyses:

```{r findOverlaps}
# Create second set of ranges
gr2 <- GRanges(
    seqnames = "chr1",
    ranges = IRanges(start = c(120, 280), end = c(180, 320)),
    strand = "*"
)

gr2_path <- tempfile(fileext = ".parquet")
gr2_df <- as.data.frame(gr2)
gr2_df$range_id <- paste0("range", 1:nrow(gr2_df))
arrow::write_parquet(gr2_df, gr2_path)
gr2_ddb <- DuckDBGRanges(gr2_path,
                         seqnames = "seqnames",
                         start = "start",
                         end = "end",
                         strand = "strand",
                         keycol = list(range_id = gr2_df$range_id))

# Find overlaps
hits <- findOverlaps(gr_ddb, gr2_ddb)
hits
```

This translates to an efficient SQL JOIN with range predicates.

### countOverlaps()

```{r countOverlaps}
# Count overlaps for each range
counts <- countOverlaps(gr_ddb, gr2_ddb)
counts
```

### overlapsAny()

```{r overlapsAny}
# Check if each range has any overlap
any_overlap <- overlapsAny(gr_ddb, gr2_ddb)
any_overlap
```

### subsetByOverlaps()

```{r subsetByOverlaps}
# Subset to ranges with overlaps
overlapping <- subsetByOverlaps(gr_ddb, gr2_ddb)
overlapping
```


## Distance Operations

### distance()

```{r distance}
# Distance between corresponding ranges
dist <- distance(gr_ddb[1:2], gr2_ddb[1:2])
dist
```

### distanceToNearest()

```{r distanceToNearest}
# Find nearest range and distance
nearest <- distanceToNearest(gr_ddb, gr2_ddb)
nearest
```


## Set Operations

### union()

```{r union}
# Union of ranges (like reduce on combined)
union_gr <- union(gr_ddb[1:2], gr_ddb[3:4])
union_gr
```

### intersect()

```{r intersect}
# Intersection of overlapping ranges
intersect_gr <- intersect(gr_ddb[1:2], gr_ddb[2:3])
intersect_gr
```

### setdiff()

```{r setdiff}
# Set difference
setdiff_gr <- setdiff(gr_ddb[1:3], gr_ddb[2:4])
setdiff_gr
```


# The DuckDBGRangesList Class

## Class Structure

*DuckDBGRangesList* extends *GRangesList* and stores grouped ranges using
DuckDB's LIST[] columns:

```{r DuckDBGRangesList-class}
showClass("DuckDBGRangesList")
```

### Slots

**frame** (*DuckDBDataFrame*)
: A DuckDBDataFrame where genomic coordinate columns (seqnames, start, end,
  width, strand) are LIST[] columns. Each row represents a group.

**seqinfo** (*Seqinfo*)
: Sequence information shared across all groups.

**elementMetadata** (*DuckDBDataFrame* or NULL)
: Per-group metadata (not per-range metadata).


## Construction

### From Parquet with LIST[] Columns

```{r DuckDBGRangesList-construct}
# Create grouped ranges data
grl_data <- data.frame(
    transcript_id = c("ENST001", "ENST002", "ENST003"),
    gene_id = c("GENE1", "GENE2", "GENE1"),
    seqnames = I(list(
        rep("chr1", 3),
        rep("chr2", 2),
        c("chr1", "chr1")
    )),
    start = I(list(
        c(100L, 200L, 300L),
        c(150L, 250L),
        c(400L, 500L)
    )),
    width = I(list(
        c(50L, 75L, 60L),
        c(80L, 45L),
        c(55L, 70L)
    )),
    strand = I(list(
        c("+", "-", "+"),
        c("-", "+"),
        c("+", "+")
    )),
    score = c(0.8, 0.6, 0.9)
)

grl_path <- tempfile(fileext = ".parquet")
arrow::write_parquet(grl_data, grl_path)

# Create DuckDBGRangesList
grl_ddb <- DuckDBGRangesList(grl_path,
                             seqnames = "seqnames",
                             start = "start",
                             width = "width",
                             strand = "strand",
                             mcols = c("gene_id", "score"),
                             keycol = list(transcript_id = grl_data$transcript_id))
grl_ddb
```


## List Operations

*DuckDBGRangesList* supports all GRangesList operations:

### Element Access

```{r DuckDBGRangesList-access}
# Number of groups
length(grl_ddb)

# Number of ranges per group
elementNROWS(grl_ddb)

# Extract single group (returns GRanges)
grl_ddb[[1]]

# Extract multiple groups (returns GRangesList)
grl_ddb[1:2]
```

### Metadata

```{r DuckDBGRangesList-metadata}
# Per-group metadata
mcols(grl_ddb)

# Group names
names(grl_ddb)
```


# SQL Translation Details

## How Range Operations Become SQL

Understanding the SQL translation helps optimize queries:

### Simple Filtering

```{r sql-filtering, eval=FALSE}
# R code
gr_chr1 <- gr_ddb[seqnames(gr_ddb) == "chr1"]

# Becomes SQL
# SELECT * FROM ranges WHERE seqnames = 'chr1'
```

### Range Predicates

```{r sql-range-predicates, eval=FALSE}
# R code
wide_ranges <- gr_ddb[width(gr_ddb) > 100]

# Becomes SQL
# SELECT * FROM ranges WHERE (end - start + 1) > 100
```

### Overlap Queries

```{r sql-overlaps, eval=FALSE}
# R code
hits <- findOverlaps(gr_ddb, gr2_ddb)

# Becomes SQL
# SELECT a.row_number, b.row_number
# FROM ranges_a a
# INNER JOIN ranges_b b
#   ON a.seqnames = b.seqnames
#   AND a.start <= b.end
#   AND a.end >= b.start
# WHERE (a.strand = b.strand OR a.strand = '*' OR b.strand = '*')
```


## Optimization Strategies

### Use Parquet Statistics

Parquet files store min/max statistics per column chunk. DuckDB uses these
for predicate pushdown:

```{r parquet-stats, eval=FALSE}
# This query only reads relevant chunks
chr22_ranges <- gr_ddb[seqnames(gr_ddb) == "chr22"]
# DuckDB skips chunks where seqnames != "chr22"
```

### Leverage Partitioning

For very large datasets, partition by chromosome:

```{r partitioning, eval=FALSE}
# Write partitioned by chromosome
arrow::write_dataset(gr_df, "ranges_partitioned",
                     partitioning = "seqnames")

# Queries on single chromosome are very fast
gr_chr1 <- DuckDBGRanges("ranges_partitioned", ...)
gr_chr1_subset <- gr_chr1[seqnames(gr_chr1) == "chr1"]
# Only reads chr1 partition
```

### Filter Before Expensive Operations

```{r filter-early, eval=FALSE}
# Good: Filter first
gr_subset <- gr_ddb[seqnames(gr_ddb) == "chr1"]
hits <- findOverlaps(gr_subset, gr2_ddb)

# Less efficient: Find overlaps on full dataset
hits_all <- findOverlaps(gr_ddb, gr2_ddb)
hits_chr1 <- hits_all[seqnames(gr_ddb)[queryHits(hits_all)] == "chr1", ]
```


# Advanced Topics

## Seqinfo Management

*DuckDBGRanges* fully supports Seqinfo:

```{r seqinfo}
# Get seqinfo
seqinfo(gr_ddb)

# Set sequence lengths
seqlengths(gr_ddb) <- c(chr1 = 248956422, chr2 = 242193529)
seqinfo(gr_ddb)

# Set genome
genome(gr_ddb) <- "hg38"
seqinfo(gr_ddb)
```


## Strand-Specific Operations

Many operations respect strand:

```{r strand-ops}
# Upstream flanks respect strand
upstream <- flank(gr_ddb, width = 50, start = TRUE)

# Positive strand: flank is to the left
# Negative strand: flank is to the right
# Unstranded: flank is to the left
```


## Metadata Propagation

Metadata columns propagate through operations:

```{r metadata-propagation}
# Metadata preserved in subsetting
subset_gr <- gr_ddb[1:3]
mcols(subset_gr)

# Metadata preserved in range operations
shifted_gr <- shift(gr_ddb, 100)
mcols(shifted_gr)
```


# Use Cases

## Variant Analysis

Large variant databases (gnomAD, ClinVar) benefit from DuckDB backend:

```{r use-case-variants, eval=FALSE}
# Load gnomAD variants (hundreds of millions)
gnomad <- DuckDBGRanges("gnomad.parquet",
                        seqnames = "chrom",
                        start = "pos",
                        end = "pos",  # SNVs have width = 1
                        strand = "*",
                        mcols = c("ref", "alt", "af", "an", "ac"))

# Find variants in gene region
gene_region <- GRanges("chr1:100000-200000")
gene_variants <- subsetByOverlaps(gnomad, gene_region)

# Filter by allele frequency
common_variants <- gene_variants[gene_variants$af > 0.01]
```


## Gene Annotation

Comprehensive gene models with millions of exons:

```{r use-case-genes, eval=FALSE}
# Load GENCODE annotations
gencode <- DuckDBGRangesList("gencode.parquet",
                             seqnames = "seqnames",
                             start = "start",
                             end = "end",
                             strand = "strand",
                             mcols = c("gene_id", "gene_name", "gene_type"),
                             keycol = "transcript_id")

# Find genes overlapping region
region <- GRanges("chr1:1000000-2000000")
overlapping_genes <- subsetByOverlaps(gencode, region)

# Get protein-coding genes
protein_coding <- gencode[gencode$gene_type == "protein_coding"]
```


## scATAC-seq Peak Analysis

Single-cell chromatin accessibility with hundreds of thousands of peaks:

```{r use-case-scatac, eval=FALSE}
# Load peak calls
peaks <- DuckDBGRanges("peaks.parquet",
                       seqnames = "chrom",
                       start = "start",
                       end = "end",
                       strand = "*",
                       mcols = c("peak_id", "score", "pvalue"))

# Find peaks near genes
tss <- promoters(gencode, upstream = 0, downstream = 1)
tss_peaks <- subsetByOverlaps(peaks, tss)

# Annotate peaks with nearest gene
nearest_genes <- nearest(peaks, tss)
```


## Regulatory Element Analysis

Genome-wide regulatory annotations:

```{r use-case-regulatory, eval=FALSE}
# Load ENCODE regulatory elements
encode <- DuckDBGRanges("encode_elements.parquet",
                        seqnames = "chrom",
                        start = "start",
                        end = "end",
                        strand = "*",
                        mcols = c("element_type", "cell_type", "score"))

# Find enhancers near gene
gene_region <- GRanges("chr1:1000000-2000000")
nearby_enhancers <- encode[encode$element_type == "enhancer" &
                           overlapsAny(encode, gene_region + 1e6)]
```


# Performance Considerations

## When to Use DuckDBGRanges

**Advantages**:
- Genomic datasets larger than RAM
- Millions of ranges (variants, peaks, annotations)
- Frequent filtering and subsetting
- SQL-based range queries
- Parquet files already on disk

**Limitations**:
- Random access slower than in-memory
- Some operations require materialization
- Overhead for small datasets (< 100K ranges)


## Optimization Tips

1. **Partition by chromosome**: Enables efficient per-chromosome queries
2. **Use Parquet statistics**: Filter predicates leverage min/max stats
3. **Filter early**: Reduce dataset size before expensive operations
4. **Leverage strand**: Strand-specific queries are more efficient
5. **Batch operations**: Multiple operations on same data reuse queries


## Benchmarking

```{r benchmark, eval=FALSE}
library(microbenchmark)

# Small dataset (in-memory faster)
small_gr <- gr_ddb[1:1000]
small_mem <- as(small_gr, "GRanges")

microbenchmark(
    duckdb = findOverlaps(small_gr, small_gr),
    memory = findOverlaps(small_mem, small_mem),
    times = 10
)

# Large dataset (DuckDB faster)
large_gr <- DuckDBGRanges("large_variants.parquet", ...)
# large_mem <- as(large_gr, "GRanges")  # Won't fit in RAM!

system.time(findOverlaps(large_gr, large_gr))  # Efficient SQL JOIN
```


# Session Information

```{r sessioninfo}
sessionInfo()
```
