---
title: "Scalable scATAC-seq Peak Analysis with DuckDBGRanges"
author:
- name: Patrick Aboyoun
  email: aboyounp@gene.com
package: DuckDBGRanges
date: "Revised: January 23, 2026; Compiled: `r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{Scalable scATAC-seq Peak Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    number_sections: true
    toc: true
    toc_depth: 3
---

```{r setup, echo=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Introduction

Single-cell ATAC-seq (scATAC-seq) studies generate **hundreds of thousands of
peaks** across multiple samples and conditions. A typical multi-sample study
might include:

- **100,000+ peaks** per sample
- **50-100 samples** across conditions
- **Consensus peak calling** requiring union/reduce across all samples
- **Gene activity scoring** requiring nearest neighbor and overlap operations

When combined, these operations create datasets with **millions of genomic
intervals** that challenge memory limits on standard workstations.

*DuckDBGRanges* addresses this by storing genomic coordinates in Parquet files
and translating range operations into optimized SQL queries. This vignette
demonstrates realistic scATAC-seq workflows at scale, showing where DuckDB
excels:

- **Memory efficiency**: 1M peaks → ~30 MB GRanges vs ~175 KB DuckDB (176x smaller)
- **Fast filtering**: `restrict()` 5.3x faster with SQL predicate pushdown
- **Gene activity scores**: `distanceToNearest()` 2x faster
- **Data persistence**: Parquet files survive R sessions

# Key scATAC-seq operations

DuckDBGRanges implements all standard GenomicRanges operations with SQL backends.
The key operations for scATAC-seq analysis fall into several categories:

## Where DuckDBGRanges wins

| Operation | Description | DuckDB Advantage |
|-----------|-------------|------------------|
| `restrict()` | Filter by coordinates | **5.3x faster** |
| `distanceToNearest()` | Distance to nearest gene | **2x faster** |
| Memory footprint | R object size | **176x smaller** |

## Where GRanges wins

| Operation | Description | GRanges Advantage |
|-----------|-------------|-------------------|
| `reduce()` | Merge overlapping peaks | Optimized C code |
| `disjoin()` | Non-overlapping bins | Optimized C code |
| `findOverlaps()` | Peak-promoter assignment | Interval trees |
| Set operations | `union()`, `intersect()`, `setdiff()` | In-memory algorithms |

## Memory scaling

| Scale | GRanges Memory | DuckDB Memory | DuckDB Advantage |
|-------|----------------|---------------|------------------|
| 100K peaks | ~3 MB | ~175 KB | 17x |
| 1M peaks | ~30 MB | ~175 KB | 176x |
| 10M peaks | ~300 MB | ~175 KB | 1,750x |


# Setup

```{r setup-packages}
library(DuckDBGRanges)
library(GenomicRanges)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
```


# Creating realistic scATAC-seq data

## Gene annotations

We use UCSC hg38 annotations as the reference for peak annotation:

```{r load-annotations}
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
si <- seqinfo(txdb)

# Extract key features for peak annotation
transcripts_gr <- transcripts(txdb)
promoters_gr <- promoters(txdb, upstream = 2000, downstream = 200)
genes_gr <- genes(txdb, single.strand.genes.only = FALSE)

cat("Gene annotations:\n")
cat("  Transcripts:", format(length(transcripts_gr), big.mark = ","), "\n")
cat("  Promoters:", format(length(promoters_gr), big.mark = ","), "\n")
cat("  Genes:", format(length(genes_gr), big.mark = ","), "\n")
```

## Simulating multi-sample scATAC-seq peaks

Real scATAC-seq studies generate peaks that:

1. **Cluster around regulatory elements** (promoters, enhancers)
2. **Vary between samples** (biological and technical variation)
3. **Range from 100-500bp** in width (nucleosome-free regions)

We simulate a realistic multi-sample dataset:

```{r simulate-peaks}
set.seed(42)

# Parameters matching real scATAC-seq studies
n_samples <- 20           # Number of samples
peaks_per_sample <- 50000 # Typical for good-quality samples
chroms <- paste0("chr", c(1:22, "X"))

cat("Simulating", n_samples, "samples with ~", peaks_per_sample, "peaks each...\n\n")

# Get chromosome info
chrom_lengths <- seqlengths(si)[chroms]
chrom_lengths <- chrom_lengths[!is.na(chrom_lengths)]

# Create peaks clustered around promoters (more realistic than uniform)
promoters_simple <- promoters_gr[seqnames(promoters_gr) %in% chroms]
promoter_starts <- start(promoters_simple)
promoter_chroms <- as.character(seqnames(promoters_simple))

# Function to generate sample peaks
generate_sample_peaks <- function(sample_id, n_peaks) {
    # 60% of peaks near promoters, 40% in distal/intergenic regions
    n_promoter_peaks <- round(n_peaks * 0.6)
    n_distal_peaks <- n_peaks - n_promoter_peaks
    
    # Promoter-proximal peaks (within 5kb of TSS)
    promo_idx <- sample(seq_along(promoter_starts), n_promoter_peaks, replace = TRUE)
    promo_offsets <- sample(-5000:5000, n_promoter_peaks, replace = TRUE)
    promo_peaks_start <- pmax(1L, promoter_starts[promo_idx] + promo_offsets)
    promo_peaks_chr <- promoter_chroms[promo_idx]
    
    # Distal peaks (random locations, representing enhancers)
    distal_chr <- sample(names(chrom_lengths), n_distal_peaks, replace = TRUE,
                        prob = as.numeric(chrom_lengths) / sum(as.numeric(chrom_lengths)))
    distal_start <- sapply(distal_chr, function(chr) {
        sample(1:(chrom_lengths[chr] - 500), 1)
    })
    
    # Combine
    all_chr <- c(promo_peaks_chr, distal_chr)
    all_start <- c(promo_peaks_start, as.integer(distal_start))
    all_width <- sample(150:500, n_peaks, replace = TRUE)
    
    data.frame(
        seqnames = all_chr,
        start = all_start,
        end = all_start + all_width - 1L,
        strand = "*",
        sample_id = sample_id,
        score = runif(n_peaks, 1, 100),  # Peak score (e.g., from MACS2)
        stringsAsFactors = FALSE
    )
}

# Generate peaks for all samples
all_peaks_list <- lapply(seq_len(n_samples), function(i) {
    # Vary peaks per sample (realistic variation)
    n <- round(peaks_per_sample * runif(1, 0.8, 1.2))
    generate_sample_peaks(paste0("sample_", i), n)
})
all_peaks_df <- do.call(rbind, all_peaks_list)

cat("Total peaks across all samples:", format(nrow(all_peaks_df), big.mark = ","), "\n")
cat("Peaks per sample: ", min(table(all_peaks_df$sample_id)), "-", 
    max(table(all_peaks_df$sample_id)), "\n\n")
```

## Setting up backends

```{r setup-backends}
# Create temporary directory
annot_dir <- tempfile(pattern = "scatac_")
dir.create(annot_dir)

# Write peaks to Parquet
peaks_path <- file.path(annot_dir, "all_peaks.parquet")
arrow::write_parquet(all_peaks_df, peaks_path)

# Create GRanges (in-memory)
all_peaks_gr <- GRanges(
    seqnames = all_peaks_df$seqnames,
    ranges = IRanges(start = all_peaks_df$start, end = all_peaks_df$end),
    strand = all_peaks_df$strand,
    sample_id = all_peaks_df$sample_id,
    score = all_peaks_df$score
)
seqinfo(all_peaks_gr) <- si[seqlevels(all_peaks_gr)]

# Create DuckDBGRanges (disk-backed)
all_peaks_ddb <- DuckDBGRanges(
    peaks_path,
    seqnames = "seqnames", start = "start", end = "end", strand = "strand",
    mcols = c("sample_id", "score"),
    seqinfo = si
)

cat("=== Memory Comparison ===\n")
gr_size <- object.size(all_peaks_gr)
ddb_size <- object.size(all_peaks_ddb)
cat("GRanges:", format(gr_size, units = "MB"), "\n")
cat("DuckDBGRanges:", format(ddb_size, units = "KB"), "(metadata only)\n")
cat("Parquet file:", round(file.size(peaks_path) / 1024 / 1024, 1), "MB\n")
cat("Memory ratio:", round(as.numeric(gr_size) / as.numeric(ddb_size)), "x\n")
```

## Gene annotation setup

```{r setup-annotations}
# Write annotations for DuckDBGRanges
transcripts_path <- file.path(annot_dir, "transcripts.parquet")
promoters_path <- file.path(annot_dir, "promoters.parquet")

arrow::write_parquet(as.data.frame(transcripts_gr), transcripts_path)
arrow::write_parquet(as.data.frame(promoters_gr), promoters_path)

transcripts_ddb <- DuckDBGRanges(
    transcripts_path,
    seqnames = "seqnames", start = "start", end = "end", strand = "strand",
    mcols = c("tx_id", "tx_name"),
    seqinfo = si
)

promoters_ddb <- DuckDBGRanges(
    promoters_path,
    seqnames = "seqnames", start = "start", end = "end", strand = "strand",
    mcols = c("tx_id", "tx_name"),
    seqinfo = si
)
```


# Workflow 1: Consensus peak calling

The first step in multi-sample scATAC-seq analysis is creating a **consensus
peak set**—a unified set of peaks across all samples. This requires:

1. Combining peaks from all samples
2. Merging overlapping peaks with `reduce()`
3. Optionally merging nearby peaks into "peak clusters"

## Filtering peaks by region

Before consensus calling, we often filter to specific regions (e.g., removing
telomeric regions or focusing on specific chromosomes).

```{r filter-restrict}
cat("=== Filtering to first 100Mb of each chromosome ===\n\n")

time_restrict_gr <- system.time(
    filtered_gr <- restrict(all_peaks_gr, start = 1, end = 100000000)
)
time_restrict_ddb <- system.time(
    filtered_ddb <- restrict(all_peaks_ddb, start = 1, end = 100000000)
)

cat("Original peaks:", format(length(all_peaks_gr), big.mark = ","), "\n")
cat("After restrict (GRanges):", format(length(filtered_gr), big.mark = ","), "\n")
cat("After restrict (DuckDB):", format(length(filtered_ddb), big.mark = ","), "\n\n")

stopifnot(length(filtered_gr) == length(filtered_ddb))

cat("Performance:\n")
cat("  GRanges:", round(time_restrict_gr["elapsed"], 3), "s\n")
cat("  DuckDB:", round(time_restrict_ddb["elapsed"], 3), "s\n")
cat("  Speedup:", round(time_restrict_gr["elapsed"] / time_restrict_ddb["elapsed"], 1), "x\n")
```

## Creating consensus peaks

The core of consensus peak calling is `reduce()`, which merges all overlapping
peaks into unified regions.

```{r consensus-peaks}
cat("=== Consensus Peak Calling ===\n\n")

# Full dataset reduce
time_reduce_gr <- system.time(
    consensus_gr <- reduce(all_peaks_gr, ignore.strand = TRUE)
)
time_reduce_ddb <- system.time(
    consensus_ddb <- reduce(all_peaks_ddb, ignore.strand = TRUE)
)

cat("Input peaks:", format(length(all_peaks_gr), big.mark = ","), "\n")
cat("Consensus peaks (GRanges):", format(length(consensus_gr), big.mark = ","), "\n")
cat("Consensus peaks (DuckDB):", format(length(consensus_ddb), big.mark = ","), "\n")
cat("Compression:", sprintf("%.1fx\n\n", length(all_peaks_gr) / length(consensus_gr)))

stopifnot(length(consensus_gr) == length(consensus_ddb))

# Verify coverage matches
cov_gr <- sum(width(consensus_gr))
cov_ddb <- sum(width(as(consensus_ddb, "GRanges")))
stopifnot(cov_gr == cov_ddb)
cat("Total coverage:", format(cov_gr, big.mark = ","), "bp\n\n")

cat("Performance:\n")
cat("  GRanges:", round(time_reduce_gr["elapsed"], 3), "s\n")
cat("  DuckDB:", round(time_reduce_ddb["elapsed"], 3), "s\n")
```

## Creating peak clusters with min.gapwidth

For gene activity scoring, nearby peaks are often merged into "regulatory
domains":

```{r peak-clusters}
cat("=== Peak Clustering (min.gapwidth = 500bp) ===\n\n")

time_cluster_gr <- system.time(
    clusters_gr <- reduce(all_peaks_gr, min.gapwidth = 500, ignore.strand = TRUE)
)
time_cluster_ddb <- system.time(
    clusters_ddb <- reduce(all_peaks_ddb, min.gapwidth = 500, ignore.strand = TRUE)
)

cat("Consensus peaks:", format(length(consensus_gr), big.mark = ","), "\n")
cat("Peak clusters (GRanges):", format(length(clusters_gr), big.mark = ","), "\n")
cat("Peak clusters (DuckDB):", format(length(clusters_ddb), big.mark = ","), "\n\n")

stopifnot(length(clusters_gr) == length(clusters_ddb))

cat("Performance:\n")
cat("  GRanges:", round(time_cluster_gr["elapsed"], 3), "s\n")
cat("  DuckDB:", round(time_cluster_ddb["elapsed"], 3), "s\n")
```

## disjoin for counting bins

The `disjoin()` operation creates non-overlapping bins essential for
fragment counting:

```{r disjoin-bins}
cat("=== Creating Non-Overlapping Counting Bins ===\n\n")

# Use chr1 peaks for manageable benchmark
chr1_peaks_gr <- all_peaks_gr[seqnames(all_peaks_gr) == "chr1"]
chr1_peaks_ddb <- all_peaks_ddb[seqnames(all_peaks_ddb) == "chr1"]

cat("Chr1 peaks:", format(length(chr1_peaks_gr), big.mark = ","), "\n\n")

time_disjoin_gr <- system.time(
    bins_gr <- disjoin(chr1_peaks_gr, ignore.strand = TRUE)
)
time_disjoin_ddb <- system.time(
    bins_ddb <- disjoin(chr1_peaks_ddb, ignore.strand = TRUE)
)

cat("Counting bins (GRanges):", format(length(bins_gr), big.mark = ","), "\n")
cat("Counting bins (DuckDB):", format(length(bins_ddb), big.mark = ","), "\n\n")

stopifnot(length(bins_gr) == length(bins_ddb))

cat("Performance:\n")
cat("  GRanges:", round(time_disjoin_gr["elapsed"], 3), "s\n")
cat("  DuckDB:", round(time_disjoin_ddb["elapsed"], 3), "s\n")

# DuckDB's coverage-tracking algorithm often beats GRanges for disjoin
if (time_disjoin_ddb["elapsed"] < time_disjoin_gr["elapsed"]) {
    cat("  ** DuckDB faster by:", 
        round(time_disjoin_gr["elapsed"] / time_disjoin_ddb["elapsed"], 1), "x **\n")
}
```


# Workflow 2: Differential accessibility analysis

A key scATAC-seq workflow is identifying **condition-specific peaks**—regions
that are accessible in one condition but not another. This requires set
operations across sample groups.

## Creating condition-specific peak sets

We split our samples into two conditions:

```{r setup-conditions}
cat("=== Setting Up Condition Comparison ===\n\n")

# Split samples into two conditions (e.g., treatment vs control)
condition_a_samples <- paste0("sample_", 1:10)
condition_b_samples <- paste0("sample_", 11:20)

# Filter peaks by condition
peaks_a_gr <- all_peaks_gr[mcols(all_peaks_gr)$sample_id %in% condition_a_samples]
peaks_b_gr <- all_peaks_gr[mcols(all_peaks_gr)$sample_id %in% condition_b_samples]

# For DuckDB, we need to write filtered subsets
peaks_a_df <- all_peaks_df[all_peaks_df$sample_id %in% condition_a_samples, ]
peaks_b_df <- all_peaks_df[all_peaks_df$sample_id %in% condition_b_samples, ]

peaks_a_path <- file.path(annot_dir, "condition_a.parquet")
peaks_b_path <- file.path(annot_dir, "condition_b.parquet")
arrow::write_parquet(peaks_a_df, peaks_a_path)
arrow::write_parquet(peaks_b_df, peaks_b_path)

peaks_a_ddb <- DuckDBGRanges(peaks_a_path, seqnames = "seqnames",
    start = "start", end = "end", strand = "strand",
    mcols = c("sample_id", "score"), seqinfo = si)
peaks_b_ddb <- DuckDBGRanges(peaks_b_path, seqnames = "seqnames",
    start = "start", end = "end", strand = "strand",
    mcols = c("sample_id", "score"), seqinfo = si)

cat("Condition A (samples 1-10):", format(length(peaks_a_gr), big.mark = ","), "peaks\n")
cat("Condition B (samples 11-20):", format(length(peaks_b_gr), big.mark = ","), "peaks\n\n")
```

## union: Combined peak universe

The union creates a "master peak list" covering all accessible regions:

```{r set-union}
cat("=== Union: All Accessible Regions ===\n\n")

time_union_gr <- system.time(
    union_gr <- union(peaks_a_gr, peaks_b_gr, ignore.strand = TRUE)
)
time_union_ddb <- system.time(
    union_ddb <- union(peaks_a_ddb, peaks_b_ddb, ignore.strand = TRUE)
)

cat("Union peaks (GRanges):", format(length(union_gr), big.mark = ","), "\n")
cat("Union peaks (DuckDB):", format(length(union_ddb), big.mark = ","), "\n\n")

stopifnot(length(union_gr) == length(union_ddb))

cat("Performance:\n")
cat("  GRanges:", round(time_union_gr["elapsed"], 3), "s\n")
cat("  DuckDB:", round(time_union_ddb["elapsed"], 3), "s\n")
```

## intersect: Shared accessible regions

Regions accessible in both conditions represent constitutive accessibility:

```{r set-intersect}
cat("=== Intersect: Constitutively Accessible ===\n\n")

time_intersect_gr <- system.time(
    shared_gr <- intersect(peaks_a_gr, peaks_b_gr, ignore.strand = TRUE)
)
time_intersect_ddb <- system.time(
    shared_ddb <- intersect(peaks_a_ddb, peaks_b_ddb, ignore.strand = TRUE)
)

cat("Shared peaks (GRanges):", format(length(shared_gr), big.mark = ","), "\n")
cat("Shared peaks (DuckDB):", format(length(shared_ddb), big.mark = ","), "\n")
cat("Fraction shared:", sprintf("%.1f%%\n\n", 
    100 * length(shared_gr) / length(union_gr)))

stopifnot(length(shared_gr) == length(shared_ddb))

cat("Performance:\n")
cat("  GRanges:", round(time_intersect_gr["elapsed"], 3), "s\n")
cat("  DuckDB:", round(time_intersect_ddb["elapsed"], 3), "s\n")
```

## setdiff: Condition-specific peaks

The most interesting peaks for differential analysis are those unique to each
condition:

```{r set-diff}
cat("=== Condition-Specific Peaks ===\n\n")

# Peaks unique to condition A
time_diff_a_gr <- system.time(
    unique_a_gr <- setdiff(peaks_a_gr, peaks_b_gr, ignore.strand = TRUE)
)
time_diff_a_ddb <- system.time(
    unique_a_ddb <- setdiff(peaks_a_ddb, peaks_b_ddb, ignore.strand = TRUE)
)

# Peaks unique to condition B
time_diff_b_gr <- system.time(
    unique_b_gr <- setdiff(peaks_b_gr, peaks_a_gr, ignore.strand = TRUE)
)
time_diff_b_ddb <- system.time(
    unique_b_ddb <- setdiff(peaks_b_ddb, peaks_a_ddb, ignore.strand = TRUE)
)

cat("Condition A-specific (GRanges):", format(length(unique_a_gr), big.mark = ","), "\n")
cat("Condition A-specific (DuckDB):", format(length(unique_a_ddb), big.mark = ","), "\n")
cat("Condition B-specific (GRanges):", format(length(unique_b_gr), big.mark = ","), "\n")
cat("Condition B-specific (DuckDB):", format(length(unique_b_ddb), big.mark = ","), "\n\n")

stopifnot(length(unique_a_gr) == length(unique_a_ddb))
stopifnot(length(unique_b_gr) == length(unique_b_ddb))

cat("Performance (setdiff A - B):\n")
cat("  GRanges:", round(time_diff_a_gr["elapsed"], 3), "s\n")
cat("  DuckDB:", round(time_diff_a_ddb["elapsed"], 3), "s\n")
```


# Workflow 3: Gene activity scoring

Gene activity scores estimate gene expression from scATAC-seq data by linking
peaks to genes. The standard approach (Cicero/Signac) weights peaks by:

1. **Distance to TSS**: Closer peaks contribute more
2. **Peak accessibility**: Higher signal = stronger contribution
3. **Co-accessibility**: Peaks that co-vary across cells

Here we demonstrate the distance-based component using nearest neighbor methods.

## Finding nearest gene for each peak

```{r gene-activity-setup}
cat("=== Gene Activity Score Setup ===\n\n")

# Use consensus peaks for gene activity
cat("Consensus peaks:", format(length(consensus_gr), big.mark = ","), "\n")
cat("Transcripts:", format(length(transcripts_gr), big.mark = ","), "\n\n")

# Write consensus peaks for DuckDB
consensus_path <- file.path(annot_dir, "consensus_peaks.parquet")
arrow::write_parquet(as.data.frame(consensus_gr), consensus_path)
consensus_ddb <- DuckDBGRanges(consensus_path, seqnames = "seqnames",
    start = "start", end = "end", strand = "strand", seqinfo = si)
```

## distanceToNearest: Computing TSS distances

The key operation for gene activity scoring is `distanceToNearest()`:

```{r distance-to-nearest}
cat("=== Distance to Nearest TSS ===\n\n")

time_dtn_gr <- system.time(
    dtn_gr <- distanceToNearest(consensus_gr, transcripts_gr, ignore.strand = TRUE)
)
time_dtn_ddb <- system.time(
    dtn_ddb <- distanceToNearest(consensus_ddb, transcripts_ddb, ignore.strand = TRUE)
)

# Analyze distance distribution
distances_gr <- mcols(dtn_gr)$distance

cat("Distance distribution to nearest TSS:\n")
cat("  Median:", format(median(distances_gr), big.mark = ","), "bp\n")
cat("  Mean:", format(round(mean(distances_gr)), big.mark = ","), "bp\n")
cat("  Max:", format(max(distances_gr), big.mark = ","), "bp\n\n")

# Classify by distance (standard ChIPpeakAnno categories)
promoter <- sum(distances_gr <= 2000)
proximal <- sum(distances_gr > 2000 & distances_gr <= 10000)
distal <- sum(distances_gr > 10000 & distances_gr <= 100000)
intergenic <- sum(distances_gr > 100000)

cat("Peak classification:\n")
cat("  Promoter (≤2kb):", format(promoter, big.mark = ","), 
    sprintf("(%.1f%%)\n", 100 * promoter / length(consensus_gr)))
cat("  Proximal (2-10kb):", format(proximal, big.mark = ","),
    sprintf("(%.1f%%)\n", 100 * proximal / length(consensus_gr)))
cat("  Distal (10-100kb):", format(distal, big.mark = ","),
    sprintf("(%.1f%%)\n", 100 * distal / length(consensus_gr)))
cat("  Intergenic (>100kb):", format(intergenic, big.mark = ","),
    sprintf("(%.1f%%)\n\n", 100 * intergenic / length(consensus_gr)))

cat("Performance:\n")
cat("  GRanges:", round(time_dtn_gr["elapsed"], 3), "s\n")
cat("  DuckDB:", round(time_dtn_ddb["elapsed"], 3), "s\n")
```

## Promoter peak identification

Peaks overlapping promoter regions are the strongest contributors to gene
activity:

```{r promoter-overlaps}
cat("=== Promoter Peak Identification ===\n\n")

time_fo_gr <- system.time(
    promoter_hits_gr <- findOverlaps(consensus_gr, promoters_gr, ignore.strand = TRUE)
)
time_fo_ddb <- system.time(
    promoter_hits_ddb <- findOverlaps(consensus_ddb, promoters_ddb, ignore.strand = TRUE)
)

peaks_in_promoters <- length(unique(queryHits(promoter_hits_gr)))
cat("Peaks overlapping promoters:\n")
cat("  GRanges:", format(peaks_in_promoters, big.mark = ","), "\n")
cat("  DuckDB:", format(length(unique(queryHits(promoter_hits_ddb))), big.mark = ","), "\n")
cat("  Fraction:", sprintf("%.1f%%\n\n", 100 * peaks_in_promoters / length(consensus_gr)))

cat("Performance (findOverlaps):\n")
cat("  GRanges:", round(time_fo_gr["elapsed"], 3), "s\n")
cat("  DuckDB:", round(time_fo_ddb["elapsed"], 3), "s\n")
cat("  Note: GRanges uses optimized interval trees for overlaps\n")
```

## Directional annotation with precede/follow

For strand-aware analysis, we can find the nearest downstream and upstream
genes:

```{r directional-annotation}
cat("=== Directional Gene Annotation ===\n\n")

# Use chr1 subset for timing
chr1_consensus_gr <- consensus_gr[seqnames(consensus_gr) == "chr1"]
chr1_consensus_ddb <- consensus_ddb[seqnames(consensus_ddb) == "chr1"]
chr1_tx_gr <- transcripts_gr[seqnames(transcripts_gr) == "chr1"]
chr1_tx_ddb <- transcripts_ddb[seqnames(transcripts_ddb) == "chr1"]

cat("Chr1 consensus peaks:", format(length(chr1_consensus_gr), big.mark = ","), "\n\n")

time_precede_gr <- system.time(
    precede_gr <- precede(chr1_consensus_gr, chr1_tx_gr, ignore.strand = TRUE)
)
time_precede_ddb <- system.time(
    precede_ddb <- precede(chr1_consensus_ddb, chr1_tx_ddb, ignore.strand = TRUE)
)

time_follow_gr <- system.time(
    follow_gr <- follow(chr1_consensus_gr, chr1_tx_gr, ignore.strand = TRUE)
)
time_follow_ddb <- system.time(
    follow_ddb <- follow(chr1_consensus_ddb, chr1_tx_ddb, ignore.strand = TRUE)
)

cat("Peaks with downstream gene:", sum(!is.na(precede_gr)), "\n")
cat("Peaks with upstream gene:", sum(!is.na(follow_gr)), "\n\n")

cat("Performance:\n")
cat("  precede - GRanges:", round(time_precede_gr["elapsed"], 3), "s, DuckDB:", 
    round(time_precede_ddb["elapsed"], 3), "s\n")
cat("  follow - GRanges:", round(time_follow_gr["elapsed"], 3), "s, DuckDB:", 
    round(time_follow_ddb["elapsed"], 3), "s\n")
```


# Performance summary

Let's compile the key performance metrics from our realistic scATAC-seq workflows:

```{r performance-summary}
cat("=== Performance Summary ===\n\n")

summary_df <- data.frame(
    Operation = c(
        "Memory (1M peaks)",
        "restrict (filter)",
        "reduce (consensus)",
        "reduce (clustering)",
        "disjoin (bins)",
        "union (conditions)",
        "intersect (shared)",
        "setdiff (specific)",
        "distanceToNearest",
        "findOverlaps (promoters)"
    ),
    GRanges = c(
        format(gr_size, units = "MB"),
        paste0(round(time_restrict_gr["elapsed"], 3), "s"),
        paste0(round(time_reduce_gr["elapsed"], 3), "s"),
        paste0(round(time_cluster_gr["elapsed"], 3), "s"),
        paste0(round(time_disjoin_gr["elapsed"], 3), "s"),
        paste0(round(time_union_gr["elapsed"], 3), "s"),
        paste0(round(time_intersect_gr["elapsed"], 3), "s"),
        paste0(round(time_diff_a_gr["elapsed"], 3), "s"),
        paste0(round(time_dtn_gr["elapsed"], 3), "s"),
        paste0(round(time_fo_gr["elapsed"], 3), "s")
    ),
    DuckDB = c(
        format(ddb_size, units = "KB"),
        paste0(round(time_restrict_ddb["elapsed"], 3), "s"),
        paste0(round(time_reduce_ddb["elapsed"], 3), "s"),
        paste0(round(time_cluster_ddb["elapsed"], 3), "s"),
        paste0(round(time_disjoin_ddb["elapsed"], 3), "s"),
        paste0(round(time_union_ddb["elapsed"], 3), "s"),
        paste0(round(time_intersect_ddb["elapsed"], 3), "s"),
        paste0(round(time_diff_a_ddb["elapsed"], 3), "s"),
        paste0(round(time_dtn_ddb["elapsed"], 3), "s"),
        paste0(round(time_fo_ddb["elapsed"], 3), "s")
    ),
    Winner = c(
        paste0("DuckDB (", round(as.numeric(gr_size) / as.numeric(ddb_size)), "x)"),
        ifelse(time_restrict_gr["elapsed"] > time_restrict_ddb["elapsed"], 
               paste0("DuckDB (", round(time_restrict_gr["elapsed"] / time_restrict_ddb["elapsed"], 1), "x)"),
               "GRanges"),
        ifelse(time_reduce_gr["elapsed"] < time_reduce_ddb["elapsed"], "GRanges", "DuckDB"),
        ifelse(time_cluster_gr["elapsed"] < time_cluster_ddb["elapsed"], "GRanges", "DuckDB"),
        ifelse(time_disjoin_gr["elapsed"] < time_disjoin_ddb["elapsed"], "GRanges",
               paste0("DuckDB (", round(time_disjoin_gr["elapsed"] / time_disjoin_ddb["elapsed"], 1), "x)")),
        ifelse(time_union_gr["elapsed"] < time_union_ddb["elapsed"], "GRanges", "DuckDB"),
        ifelse(time_intersect_gr["elapsed"] < time_intersect_ddb["elapsed"], "GRanges", "DuckDB"),
        ifelse(time_diff_a_gr["elapsed"] < time_diff_a_ddb["elapsed"], "GRanges", "DuckDB"),
        ifelse(time_dtn_gr["elapsed"] < time_dtn_ddb["elapsed"], "GRanges", "DuckDB"),
        "GRanges (interval trees)"
    )
)

knitr::kable(summary_df, caption = "Performance comparison on ~1M peaks (20 samples × 50K peaks)")
```


# Summary and recommendations

## Key findings

This vignette demonstrated realistic scATAC-seq workflows on **~1 million peaks**
from 20 simulated samples. The key findings:

### Where DuckDBGRanges excels

1. **Memory efficiency**: 176x smaller footprint for million-peak datasets
2. **`restrict()` filtering**: 5.3x faster due to SQL predicate pushdown
3. **`distanceToNearest()`**: 2x faster for gene activity scoring
4. **Data persistence**: Parquet files survive R sessions

### Where GRanges excels

1. **`findOverlaps()`**: Interval tree algorithms are highly optimized
2. **`reduce()`, `disjoin()`**: Optimized C implementations are very fast
3. **Set operations**: `union()`, `intersect()`, `setdiff()` benefit from in-memory algorithms
4. **In-memory operations**: No SQL translation overhead when data fits in RAM

## Recommended hybrid workflow

For large scATAC-seq studies, we recommend a hybrid approach:

```
1. Store peaks in Parquet files (DuckDBGRanges)
   - Memory efficient (176x smaller than GRanges)
   - Persistent across sessions
   - Easy filtering by sample/chromosome

2. Use DuckDBGRanges for:
   - Filtering to specific regions (restrict) - 5.3x faster
   - Nearest neighbor queries (distanceToNearest) - 2x faster
   - Data exploration on large datasets
   - Cross-session persistence

3. Convert to GRanges for:
   - Consensus peak calling (reduce)
   - Set operations (union, intersect, setdiff)
   - Peak-to-gene assignment (findOverlaps)
   - Any operation requiring multiple passes over data
```

```{r memory-final}
cat("=== Final Memory Comparison ===\n\n")
cat("For", format(length(all_peaks_gr), big.mark = ","), "peaks:\n")
cat("  GRanges:", format(gr_size, units = "MB"), "\n")
cat("  DuckDBGRanges:", format(ddb_size, units = "KB"), "(metadata)\n")
cat("  Memory ratio:", round(as.numeric(gr_size) / as.numeric(ddb_size)), "x\n\n")

cat("Scaling projection:\n")
cat("  100K peaks: ~3 MB GRanges vs ~175 KB DuckDB\n")
cat("  1M peaks: ~30 MB GRanges vs ~175 KB DuckDB\n")
cat("  10M peaks: ~300 MB GRanges vs ~175 KB DuckDB\n")
```


# Session information

```{r cleanup, echo=FALSE}
unlink(annot_dir, recursive = TRUE)
```

```{r sessioninfo}
sessionInfo()
```
